---
title: "RDojo 01/18/2018 - Introduction to R"
output: html_notebook
---

#0: Some Preliminaries
###0a: Working Directories

```{r}
getwd() #see wd
rm(list = ls()) #clears EVERYTHING
setwd("~/Desktop/Dropbox/blah") #set wd
```

###0b: Installing packages
If a package has not already been installed (through R) on the computer you are using, you will need to use install.packages("nameinquotes").
```{r}
install.packages("name")
```

###0c: Using installed packages
library(namenotinquotes) must be used every time you open a new session in RStudio; equivalently, you can use the "packages" section of the bottom right box to check the box next to that package once it has been installed.
```{r}
library(name)
```

###0d: Datasets included with base R
R includes a number of datasets useful for playing around and learning how R works. We will be using a number of these today. Read these datasets in using the data(X) function.
```{r}
data(iris)
data(mtcars)
```

###0e: Function information

If you don't know how to use a function, or just want to know more about how it works, you can use ?functionname __*without*__ parentheses to pull up a help box in the bottom right toolpane.
```{r}
?View
```

###0f: Function code

If you want to see the actual code that R uses when it runs a function, type the function alone, __*without*__ parentheses or a question mark.
```{r}
View
```

###0g: Non-code comments

Use # to comment out parts of code so that they don't run

```{r}
data(iris) #this is a comment
```

#1: Importing data
###1a: read.table()/read.csv() - read in general data files
read.table() and read.csv() read in general data files. 
```{r}
data1 <- read.table(file.choose())
```
-file.choose() pulls up a drop-down menu to choose the file manually, instead of using the file path.  
-The argument "sep = " tells the function what symbol is used to denote a new cell. For csv's ("comma-separated files"), this is a comma, so read.csv(file.choose()) and read.table(file.choose() , sep = ",") are equivalent in that sense.  
-The argument "header = TRUE/FALSE" tells the function to either register the first row of the dataset as a row of column names (TRUE) or not (FALSE).  
-The agument "na.rm = TRUE/FALSE" tells the function remove (TRUE) or keep (FALSE) rows that contain __*any*__ missing values.

###1b: read.spss - read in SPSS files
read.spss() reads in SPSS files. this function requires the "foreign" package.
```{r}
install.packages("foreign")
library(foreign)
data1 <- read.spss(file.choose())
```


#2: Functions for basic info about data
###2a: View()
View() shows the chosen data or object in a separate tab
```{r}
View(mtcars)
```


###2b: head()
head() shows the first 6 observations of an object
```{r}
head(mtcars)
```

###2c: str()
str() gives a summary of the structure of an object
```{r}
str(mtcars)
```

###2d: summary()
summary() gives various descriptives about all of the variables in an object
```{r}
summary(mtcars)
```

#3: dplyr package intro

###3.0: Install/load dplyr

```{r}
install.packages("dplyr")
library(dplyr)
```

###3a: filter()
filter() selects rows meeting certain conditions  
The following function creates a dataset with all observations from mtcars with 6 cylinders and 4 gears.
```{r}
cyl6.gear4 <- filter(mtcars, cyl == 6, gear == 4)
```
  
The following function creates a dataset with all observations from mtcars with 6 or more cylinders and less than 4 gears.
```{r}
cyl6p.gearLT4 <- filter(mtcars, cyl >= 6 , gear < 4) 
```

###3b: arrange()
arrange() arranges dataset by values of particular columns

```{r}
arrange(mtcars, cyl, disp) #arranges and prints by cyl, and then by disp within each level of cyl, each in ascending order
arrange(mtcars, desc(disp)) #same as above, except disp arrangement within cyl levels is in descending order
cyldisp.ord <- arrange(mtcars, cyl, disp) #creates new arranged dataset "cyldisp.ord"
```

###3c: select()
select() allows you to select particular columns of dataset without the others, while also renaming the selected columns (if desired).
```{r}
select(iris , Petal.Length) #selects and prints only the variable "Petal.Length" from the dataset, without renaming it
select(iris, petal_length = Petal.Length) #selects and prints original "Petal.Length" as "petal_length"
iris.PLength <- select(iris, petal_length = Petal.Length) #saves "Petal.Length" as "petal_length" in new object "iris.PLength"
```

###3d: rename()
rename() keeps all variables of a given dataset, bwhile renaming specified variables

```{r}
rename(iris, petal_length = Petal.Length) #renames "Petal.Length" to "petal_length" IN ORIGINAL OBJECT, while keeping remaining variables the same
iris.rename <- rename(iris, petal_length = Petal.Length) #creates new object "iris.rename" with "Petal.Length" renamed to "petal_length"
```

#4 - standard analyses

###4a: ANOVA
You can create ANOVA models using the function aov and running a summary on the the created model object.

```{r}
mpgONcyl <- aov(mtcars$mpg ~ mtcars$cyl) #creates an object with results from the ANOVA, with cyl as the group variable and mpg as the dependent variable
summary(mpgONcyl) #prints the ANOVA table summary from the created object
```

Alternatively, you can create an OLS model and then use the anova() command to run an ANOVA on the created OLS model.

```{r}
cyl.fit <- lm(mtcars$mpg ~ mtcars$cyl) #creates an OLS model with mpg as the DV and cyl as the IV
anova(cyl.fit) #runs an ANOVA on the OLS model created - notice that these results are the same as the results from summary(mpgONcyl)
```

###4b: linear model (OLS)
The lm() function creates an OLS model, with your dependent variable on the left of the tilde, and your predictor variable or variables on the right of the tilde.

```{r}
fit1 <- lm(mtcars$mpg ~ mtcars$wt) #runs a regression of mpg on wt
summary(fit1) #gives results of regression model
fit1.b <- lm(mpg ~ wt , data = mtcars) #same as above, except with the "data = X" argument instead of "data$X"
```

You can also include interaction terms. Using var1\*var2 includes both the interaction and the lower-order (main-effect) terms, while var1:var2 includes __*only*__ the interaction term.

```{r}
fit2 <- lm(mtcars$mpg ~ mtcars$wt*mtcars$cyl) #regression of mpg on interaction term between wt and cyl plus singular mpg and cyl terms.
summary(fit2)
```

```{r}
fit2.b <- lm(mtcars$mpg ~ mtcars$wt:mtcars$cyl) #regression of mpg on JUST the interaction term
summary(fit2.b)
```

```{r}
fit2.c <- lm(mtcars$mpg ~ mtcars$wt + mtcars$cyl + mtcars$wt:mtcars$cyl) #equivalent to "fit2"
summary(fit2.c)
```

###4c: generalized linear models - RCT example taken from Dobson (1990), page 93
The glm(function) allows you to run more flexible generalized linear models, using more a variety of built-in and self-created distributional family forms. Here, we create an example of count data, and run a glm using a Poisson family distribution appropriate for count data.

```{r}
counts <- c(18,17,15,20,10,20,25,13,12) #creates a vector of counts of events
outcome <- gl(3,1,9) #generates a vector of factor levels, with 3 levels, 1 replication each, for a total of 9 observations (1-3, 3 times)
treatment <- gl(3,3) #generates a vector of factor levels, with 3 levels, 3 replications each
treatment.b <- gl(3,3,9) #equivalent to "treatment"
d.AD <- data.frame(treatment, outcome, counts) #creates object "d.AD" combining three created vectors
glm.D93 <- glm(d.AD$counts ~ d.AD$outcome + d.AD$treatment, family = poisson()) #creates a GLM with created dataset, using a Poisson distribution. 
```

```{r}
anova(glm.D93) #ANOVA results of glm fit
```

```{r}
summary(glm.D93) #full summary of GLM model
```

#Extras

###Extra 1 - $ for variable indication
You can use data$variable to select the *named* column "variable" in the object "data". You can also select rows (or columns) of a dataset meeting certain conditions; we will using row/column indices and brackets in a bit. The following
commands are equivalent to the commands in __Section 3a__, plus a generic version.
```{r}
cyl6.gear4 <- mtcars[mtcars$cyl == 6 & mtcars$gear == 4 , ]
cyl6p.gearLT4 <- mtcars[mtcars$cyl >= 6 & mtcars$gear < 4 , ]
dat2 <- dat1[dat1$var1 == X & dat1$var2 == Y , ]
```

###Extra 2 - index/bracket notation
You can use brackets to indicate particular portions of an object.  
If you are using a 1-dimensional object (a vector), only one number (or set of numbers) will be in the bracket. This tells you which observation (or set of observations) within the vector you are referring to.  
If you are using a 2-dimensional object with both rows and columns (such as a dataset), there will be two numbers in the brackets separated by a comma. The first number refers to the row number or numbers; the second refers to the column number or numbers.

```{r}
vec1[A] #prints Ath observation of object "vec1"
dat1[R , C] #prints observation in Rth row and Cth column of object "dat1"
```

You can use ":" to indicate a range selection.
```{r}
vec1[A:D] #prints Ath-Dth observations of object "vec1"
dat1[R1:R4 , C] #Prints R1th through R4th rows of Cth column of "dat1"
dat1[R , C1:C4] #Prints observations in Rth row and C1th through C4th column of "dat1"
```

With 2-dimensional objects, you can also select all of the rows, or all of the columns, by leaving one or the other side of the comma blank.

```{r}
dat1[ , C] #prints all rows of Cth column of "dat1"
dat1[R , ] #prints all columns of Rth row of "dat1"
```

You can use c() to indicate a non-continuous group of observations
```{r}
vec1[c(A , B , E , Z)] #prints out observations A, B, E, and Z of "vec1"
```

You can use "-" to indicate all but an observation (or set of observations)
```{r}
vec1[-A] #prints all but Ath observation of object "vec1"
vec1[-c(A , B , E , Z)] #prints all but observations A, B, E, and Z of "vec1"
```

The above functions only print out the result of you reference; you can also create a new object using this reference.

```{r}
vec2 <- vec1[-A] #creates a new 1-dimensional object, "vec2", which contains all but Ath observation of "vec1"
```

###Extra 3: Practice codes for Index/Bracket Notation
What do each of these functions do?

```{r}
vec1[c(I,J,L,X)]
```

```{r}
vec3 <- vec1[c(I , J , L , X)]
```

```{r}
dat1[R , A:D]
```

```{r}
dat3 <- dat1[R , C1:C4]
```

```{r}
dat4 <- dat1[c(R1 , R2 , R5) , ]
```

```{r}
dat5 <- dat1[-R1 , c(C5 , C7 , C9)]
```
